# Code {-}

## Chapter 2 {-}

### Code block 2.1/2.2 {-}
```{r cb02-1/2}
aScalar <- 4
```

### Code block 2.3/2.4 {-}
This is a [ggplot2](https://ggplot2.tidyverse.org/) rather than base R version. But _ggplot2_ figures do look so much better.
```{r cb02-3/4}
library(ggplot2)

v <- c(2, -1)
ggplot(data=NULL, aes(x = c(0, v[1]), y=c(0, v[2])))  + 
  geom_line() + 
  geom_point(aes(x=v[1], y=v[2])) + 
  scale_x_continuous(name=expression(paste(X[1], " dim.")), limits = c(-3, 3)) + 
  scale_y_continuous(name=expression(paste(X[2], " dim.")), limits = c(-3, 3)) + 
  coord_equal()
```

### Code block 2.5/2.6 {-}
In R _atomic_ vectors are created via `c()` function but, technically, they are neither column, nor row vectors. [Matrix multiplication manual](https://stat.ethz.ch/R-manual/R-devel/library/base/html/matmult.html) states that it "_multiplies two matrices, if they are conformable. If one argument is a vector, it will be promoted to either a row or column matrix to make the two arguments conformable. If both are vectors of the same length, it will return the inner product (as a matrix)._" At the same time, _transposing_ an atomic vector `t(c(...))` transforms it into a single row matrix, thus a row vector, hinting that deep down an outcome of `c()` is a column vector.

To avoid ambiguity, I will use single row and single column matrices for, respectively, row and column vectors.

```{r cb02-5/6}
v1 <- matrix(c(2, 5, 5, 7), nrow = 1) # row vector
v2 <- matrix(c(2, 5, 5, 7), ncol = 1) # column vector
v3 <- matrix(c(2, 5, 5, 7))           # also a column vector
```

### Code block 2.7/2.8 {-}
```{r cb02-7/8}
v1 <- matrix(c(2, 5, 5, 7), nrow = 1) # row vector
v1a <- t(c(2, 5, 5, 7))               # also a row vector
v2 <- t(v1)                           # column vector
```
 
### Code block 2.9/2.10 {-}
For this example, you can also use atomic vectors directly without turning them into a column vector / single column matrix.
```{r cb02-9/10}
v1 <- matrix(c(2, 5, 4, 7), ncol=1)
v2 <- matrix(c(4, 1, 0, 2), ncol=1)
v3 <- 4 * v1  - 2 * v2
```

## Chapter 3 {-}

### Code block 3.1/3.2 {-}
There is no _explicit_ dot product function in base R but there are multiple implementations in various libraries such as [pracma](https://github.com/cran/pracma) used here.
```{r cb03-1/2-1}
v1 <- matrix(c(2, 5, 4, 7), ncol=1)
v2 <- matrix(c(4, 1, 0, 2), ncol=1)
dp <- pracma::dot(v1, v2)
```

However, a matrix multiplication of _atomic_ vectors (we can convert a matrix back to an atomic vector via `c()` or `as.vector()`) gives us the dot product (well, the _inner_ product, which is why the result is a 1-by-1 matrix).
```{r cb03-1/2-2}
dp <- c(v1) %*% as.vector(v2)
```

### Code block 3.3/3.4 {-}
```{r cb03-3/4}
l1 <- 1
l2 <- 2
l3 <- -3
v1 <- matrix(c(4, 5, 1), ncol=1)
v2 <- matrix(c(-4, 0, -4), ncol=1)
v3 <- matrix(c(1, 3, 2), ncol=1)
l1 * v1 + l2 * v2 + l3 * v3
```

### Code block 3.5/3.6 {-}
To compute the outer product, we must use atomic vectors, thus I've skipped the whole creating-column-vector-as-a-matrix thing.
```{r cb03-5/6-1}
v1 <- c(2, 5, 4, 7)
v2 <- c(4, 1, 0, 2)
op <- outer(v1, v2)
op <- v1 %o% v2 # alternative call as operation
```

Alternatively, if we do start with vectors as single column matrices...
```{r cb03-5/6-2}
v1 <- matrix(c(2, 5, 4, 7), ncol=1)
v2 <- matrix(c(4, 1, 0, 2), ncol=1)
op <- outer(c(v1), c(v2))
op <- c(v1) %o% c(v2) # alternative call as operation
```

### Code block 3.7/3.8 {-}
```{r cb03-7/8}
v1 <- matrix(c(2, 5, 4, 7), ncol=1)
v2 <- matrix(c(4, 1, 0, 2), ncol=1)
v3 <- v1 * v2
```

### Code block 3.9/3.10 {-}
Please note that you need to explicitly specify the 2-norm via `type="2"`, as the _one norm_ is used by default.
```{r cb03-9/10}
v <- matrix(c(2, 5, 4, 7), ncol=1)
vMag <- norm(v, type="2")
v_unit <- v / vMag
```

## Chapter 5 {-}

### Code block 5.1/5.2 {-}
There is only one way to transpose a matrix in R: via function [t()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/t.html).

```{r cb05-1/2}
A <- matrix(runif(n=2*5), nrow=2, ncol=5)
At1 <- t(A)
```

### Code block 5.3/5.4 {-}
```{r cb05-3/4}
I <- diag(4)
O <- matrix(1, nrow=4, ncol=1) # or rep(1, times=4) to create an atomic vector
Z <- matrix(0, nrow=4, ncol=4)
```

### Code block 5.5/5.6 {-}
```{r cb05-5/6}
D <- diag(c(1, 2, 3, 5))  # diagonal matrix
R <- matrix(runif(n = 3 * 4), nrow=3, ncol=4)
d <- diag(R) # diagonal elements
```

### Code block 5.7/5.8 {-}
In r [cbind()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/cbind.html) and [rbind()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/cbind.html) concatenate, respectively, by column and row.
```{r cb05-7/8}
A <- matrix(runif(n = 3 * 5), nrow=3, ncol=5)
B <- matrix(runif(n = 3 * 4), nrow=3, ncol=4)
AB <- cbind(A, B)
```

### Code block 5.9/5.10 {-}
There are two ways to compute lower and upper triangular parts of a matrix. First, to use `tril()` and `triu()` function from _pracma_ library.
```{r cb05-9/10-1}
A <- matrix(runif(n = 5 * 5), nrow=5, ncol=5)
L <- pracma::tril(A)
U <- pracma::triu(A)
```

Alternatively, you can use base R functions [lower.tri()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/lower.tri.html) and `upper.tri()` that give you a matrix of the same size with _logical_ values indicating whether an element belongs to, respectively, lower or upper triangle. Note that by default, the diagonal _is not_ included!
```{r cb05-9/10-2}
A <- matrix(runif(n = 5 * 5), nrow=5, ncol=5)

L <- A
# by setting the UPPER triangular part to 0, we get the LOWER triangular part and the diagonal
L[upper.tri(L)] <- 0

U <- A
# by setting the LOWER triangular part to 0, we get the UPPER triangular part and the diagonal
U[lower.tri(U)] <- 0
```

### Code block 5.11/5.12 {-}
Note that there is a `toeplitz()` function in _stats_ (base R) and `Topelitz()` (note the first capital letter) in _pracma_ library. Here, I use base R version.
```{r cb05-11/12}
t <- c(1, 2, 3, 4)
T <- stats::toeplitz(t)
H <- pracma::hankel(t, b= c(t[-1], t[1]))
```

### Code block 5.13/5.14 {-}
```{r cb05-13/14}
l <- 0.01
I <- diag(4)
A <- matrix(runif(4 * 4), nrow=4, ncol=4)
As <- A + l * I
```

### Code block 5.15/5.16 {-}
Base R does not implement trace function, as it is simply a `sum(diag(M))`. However, you can use `Trace()` from _pracma_ library.
```{r cb05-15/16}
A <- matrix(runif(4 * 4), nrow=4, ncol=4)
tr <- pracma::Trace(A)
```


## Chapter 6 {-}
### Code block 6.1/6.2 {-}
```{r cb06-1/2}
M1 <- matrix(runif(n=4*3), nrow=4, ncol=3)
M2 <- matrix(runif(n=3*5), nrow=3, ncol=5)
C <- M1 %*% M2
```

### Code block 6.3/6.4 {-}
```{r cb06-3/4}
A <- matrix(runif(n=2*2), nrow=2, ncol=2)
B <- matrix(runif(n=2*2), nrow=2, ncol=2)
C1 <- A %*% B
C2 <- B %*% A
```

### Code block 6.5/6.6 {-}
```{r cb06-5/6}
M1 <- matrix(runif(n=4*3), nrow=4, ncol=3)
M2 <- matrix(runif(n=4*3), nrow=4, ncol=3)
C <- M1 * M2
```


### Code block 6.7/6.8 {-}
Note that by default, matrix is constructed by column. To match the code, we need to use `byrow=TRUE` option.
```{r cb06-7/8}
A <- matrix(c(1, 2, 3, 4, 5, 6), nrow=2, byrow = TRUE)
c(A)
```

### Code block 6.9/6.10 {-}
```{r cb06-9/10}
A <- matrix(runif(n=4*3), nrow=4, ncol=3)
B <- matrix(runif(n=4*3), nrow=4, ncol=3)
f <- pracma::Trace(t(A) %*% B)
```

### Code block 6.11/6.12 {-}
```{r cb06-11/12}
A <- matrix(runif(n=4*3), nrow=4, ncol=3)
norm(A, type="F")
```

## Chapter 7 {-}

### Code block 7.1/7.2 {-}
You can use `Rank()` from the _pracma_ library. Alternatively, you can use `rankMatrix()` function from the _Matrix_ library, which in addition to the rank itself, returns information on the method used to estimate the rank via attributes. 
```{r cb07-1/2}
A <- matrix(runif(n=3*6), nrow=3, ncol=6)
r1 <- pracma::Rank(A)
r2 <- Matrix::rankMatrix(A)
```

### Code block 7.3/7.4 {-}
```{r cb07-3/4}
s <- runif(n=1)
A <- matrix(runif(n=3*5), nrow=3, ncol=5)
r1 <- pracma::Rank(A)
r2 <- pracma::Rank(s * A)
print(c(r1, r2))
```

### Code block 7.5/7.6 {-}
Source code for `Rank()` function from _pracma_ library.
```{r  cb07-5/6-1}
pracma::Rank
```

Source code for `rankMatrix()` function from _Matrix_ library.
```{r cb07-5/6-2}
Matrix::rankMatrix
```

## Chapter 8 {-}

### Code block 8.1/8.2 {-}
```{r cb08-1/2}
A <- matrix(runif(n=3*4), nrow=3, ncol=4)
pracma::nullspace(A)
```

## Chapter 9 {-}
### Code block 9.1/9.2 {-}
```{r cb09-1/2}
z <- complex(real=3, imaginary=4)
Z <- complex(length.out=2, real=0, imaginary=0)
Z[1] <- 3 + 4i
```

### Code block 9.3/9.4 {-}
Base R does not have a function that generates random integers on the interval. I will use `sample()` with replacement from the range of integers to replicate this. Also note that I have renamed `i` to `im` to minimize the confusion. 
```{r cb09-3/4}
r <- sample(-3:4, size=3, replace = TRUE)
im <- sample(-3:4, size=3, replace = TRUE)
Z <- r + im * 1i
print(Z)
print(Conj(Z))
```

### Code block 9.5/9.6 {-}
`pracma::dot()` implements the Hermitian dot product.
```{r cb09-5/6}
v <- c(0, 1i)
pracma::dot(v, v)
```

## Chapter 10 {-}
### Code block 10.1/10.2 {-}
You can use `pracma::lu()` function but note that it works only on _square_, positive definite matrices.
```{r cb10-1/2}
# Using the square matrix from practice problem b
A <- matrix(c(2, 0, 1, 1, 1, 2, 3, 1, 3), nrow=3, byrow=TRUE)
pracma::lu(A)
```

### Code block 10.3/10.4 {-}
```{r cb10-3/4}
A <- matrix(runif(n=2*4), nrow=2, ncol=4)
pracma::rref(A)
```

## Chapter 11 {-}
### Code block 11.1/11.2 {-}
```{r cb11-1/2}
A <- matrix(runif(n=3*3), nrow=3, ncol=3)
det(A)
```

## Chapter 12 {-}
### Code block 12.1/12.2 {-}
In R, you find the inverse via [solve()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/solve.html). The latter solves an equation $Ax = b$, omitting the second argument defaults `b = I` and the equation is solved to find the inverse. I have added [round()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Round.html) to make it easier to see that $A A^-1$ produces an identity matrix.
```{r cb12-1/2}
A <- matrix(rnorm(n=3*3), nrow=3, ncol=3)
Ai <- solve(A)
round(A %*% Ai, 4)
```

### Code block 12.3/12.4 {-}
```{r cb12-3/4}
A <- matrix(rnorm(n=3*3), nrow=3, ncol=3)
Acat <- cbind(A, diag(1, nrow=3, ncol=3))
Ar <- pracma::rref(Acat) # RREF
Ar <- Ar[, 4:6] # keep inverse
Ai <- solve(A)
round(Ar - Ai, 4)
```

### Code block 12.5/12.6 {-}
```{r cb12-5/6}
A <- matrix(rnorm(n=5*3), nrow=5, ncol=3)
Al <- solve(t(A) %*% A) %*% t(A)
round(Al %*% A, 4)
```
### Code block 12.7/12.8 {-}
```{r cb12-7/8}
A <- matrix(rnorm(n=3*3), nrow=3, ncol=3)
A[2, ] <- A[1, ]
Api <- pracma::pinv(A)
Api %*% A
```

## Chapter 13 {-}
### Code block 13.1/13.2 {-}
```{r cb13-1/2}
A <- matrix(c(1, 2, 3, 1, 1, 1), nrow=3, byrow=TRUE)
b <- matrix(c(5.5, -3.5, 1.5), nrow=3)
lsfit(A, b)$coefficients
```

### Code block 13.3/13.4 {-}
In R, you first perform QT decomposition via [qr()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/qr.html) and get an object. Then, you can extract component matrices of the decomposition via [qr.Q()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/qraux.html) and [qr.R()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/qraux.html).
```{r cb13-3/4}
A <- matrix(rnorm(n=4*3), nrow=4, ncol=3)
QR <- qr(A)
Q <- qr.Q(QR)
R <- qr.R(QR)
```

## Chapter 15 {-}
### Code block 15.1/15.2 {-}
```{r cb15-1/2}
A <- matrix(c(2, 3, 3, 2), nrow=2, ncol=2, byrow=TRUE)
eigenA <- eigen(A)
V <- eigenA$vectors
L <- eigenA$values
```

### Code block 15.3/15.4 {-}
```{r cb15-3/4}
n <- 3
A <- matrix(rnorm(n=n^2), nrow=n, ncol=n)
B <- matrix(rnorm(n=n^2), nrow=n, ncol=n)
eigenAB <- geigen::geigen(A, B)
evecs <- eigenAB$vectors
evals <- eigenAB$vals
```

## Chapter 16 {-}
### Code block 16.1/16.2 {-}
Note that by default number of left (matrix **U**) and right (matrix **V**) singular vectors is determined as, respectively, `nu = min(n, p)` and `nv = min(n, p)` for the _n × p_ matrix. Therefore, I included `nv=ncol(A)` to replicate output by Python. Also note that R, like Matlab, return **V**. Also note that singular values attribute is `d` and it is a vector not a diagonal matrix.
```{r cb16-1/2}
A <- matrix(c(1, 1, 0, 0, 1, 1), nrow=2, ncol=3, byrow=TRUE)
svdA <- svd(A, nv=ncol(A))
U <- svdA$u
s <- svdA$d
V <- svdA$v
```

### Code block 16.3/16.4 {-}
```{r cb16-3/4}
A <- matrix(rnorm(n=5^2), nrow=5, ncol=5)
s <- svd(A)$d
condnum <- max(s) / min(s)

#compare above with pracma::cond() 
print(c(condnum, pracma::cond(A)))
```

## Chapter 17 {-}

### Code block 17.1/17.2 {-}
```{r cb17-1/2}
m <- 4
A <- matrix(rnorm(n=m^2), nrow=m, ncol=m)
v <- matrix(rnorm(n=m), nrow=1, ncol=m)
v %*% A %*% t(v)
```
